\documentclass{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{mathpazo}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\begin{document}

\section*{Formal Specification: JQ $\to$ SQL Correctness}

This document formalizes a translation from jq-style queries over JSON records into SQL-style queries over a \emph{columnar} relational store, and proves that the translation preserves meaning. The schema has three attributes of three different types; the query language covers filtering, insertion, deletion, an aggregate, and per-row updates.

% =====================================================

\subsection*{1. Data Structures and Domains}

This section defines the two representations of the same logical data: a user database stored as JSON records and as a columnar SQL store.

\paragraph{Roles.}
A finite enum used as the third schema attribute:
\[
\text{Role} = \{\,\text{student},\ \text{employee},\ \text{retired}\,\}.
\]

\paragraph{User domains.}
Let $\mathcal{J}$ be the domain of JSON users, where each $j \in \mathcal{J}$ is a record:
\[
j = \{ \text{name} : \text{String},\ \text{age} : \mathbb{N},\ \text{role} : \text{Role} \}.
\]

Let $\mathcal{S}$ be the domain of SQL users, where each $s \in \mathcal{S}$ is a tuple in a right-nested product type so that $s.1$, $s.2.1$, and $s.2.2$ recover the three components:
\[
\mathcal{S} \;=\; \text{String} \times \mathbb{N} \times \text{Role}.
\]

\paragraph{Databases.}
\begin{itemize}
    \item \textbf{JSON Database ($\mathcal{JD}$):} A finite list of JSON user records,
    \[
        \mathcal{JD} \;=\; \text{List}\,\mathcal{J}.
    \]
    \item \textbf{SQL Database ($\mathcal{SD}$):} A \emph{columnar} record consisting of three parallel lists, one per attribute:
    \[
        SD \;=\; \big(\,\text{names} : \text{List String},\ \text{ages} : \text{List }\mathbb{N},\ \text{roles} : \text{List Role}\,\big).
    \]
    The three columns are aligned positionally: the $i$-th logical row of $SD$ is the triple $(SD.\text{names}[i],\,SD.\text{ages}[i],\,SD.\text{roles}[i])$. The columns are not assumed to have equal length; only the common prefix gives rise to logical rows.
\end{itemize}

\paragraph{Row projection.}
To bridge the columnar storage with a relational view we define a 3-way zip,
\[
\text{toRows} : \mathcal{SD} \to \text{List}\,\mathcal{S},
\qquad
\text{toRows}(SD) \;=\; \text{zip}_3\big(SD.\text{names},\ SD.\text{ages},\ SD.\text{roles}\big),
\]
which returns the list of $(\text{name}, \text{age}, \text{role})$ triples encoded by the three columns. All semantic statements about $SD$ are mediated through $\text{toRows}(SD)$.

\paragraph{Conversion functions.}
Between $\mathcal{J}$ and $\mathcal{S}$ we have a structural isomorphism:
\[
\text{toS} : \mathcal{J} \to \mathcal{S}, \qquad
\text{toJ} : \mathcal{S} \to \mathcal{J},
\]
with
\[
\text{toS}\big(\{\text{name}=n,\ \text{age}=a,\ \text{role}=r\}\big) \;=\; (n, a, r),
\]
\[
\text{toJ}\big((n,a,r)\big) \;=\; \{\text{name}=n,\ \text{age}=a,\ \text{role}=r\}.
\]
These satisfy $\text{toJ} \circ \text{toS} = \text{id}_{\mathcal{J}}$ and $\text{toS} \circ \text{toJ} = \text{id}_{\mathcal{S}}$.

% =====================================================

\subsection*{2. Database Equivalence}

The development uses two equivalence relations between $\mathcal{JD}$ and $\mathcal{SD}$, of strictly increasing strength.

\paragraph{Set equivalence.}
Two databases $JD \in \mathcal{JD}$ and $SD \in \mathcal{SD}$ are \emph{set-equivalent}, written $JD \equiv SD$, when their contents agree as multisets-modulo-multiplicity:
\begin{align*}
JD \equiv SD \iff{} & \big( \forall j \in \mathcal{J},\ j \in JD \leftrightarrow \text{toS}(j) \in \text{toRows}(SD) \big) \\
                  \land\ & \big( \forall s \in \mathcal{S},\ s \in \text{toRows}(SD) \leftrightarrow \text{toJ}(s) \in JD \big).
\end{align*}
This relation suffices for filter-based queries (\texttt{find}, \texttt{drop}), insertion (\texttt{prepend}), and emptying (\texttt{clear}).

\paragraph{Permutation equivalence.}
Two databases are \emph{permutation-equivalent}, written $JD \approx SD$, when the JSON list and the row projection of the columnar list are permutations of each other:
\[
JD \approx SD \;\iff\; (\,JD \,\textsf{.map}\, \text{toS}\,)\ \text{is a permutation of}\ \text{toRows}(SD).
\]
Permutation equivalence preserves multiplicities, which is required for the \emph{aggregate} query \texttt{count} and for any per-row \emph{update} (\texttt{modify}). Permutation equivalence implies set equivalence; the converse does not hold.

% =====================================================

\subsection*{3. Schema, Conditions, and Per-Row Update}

\paragraph{Columns and values.}
The schema is parameterised by a column tag $\text{Col} = \{\,\text{name},\ \text{age},\ \text{role},\ \text{all}\,\}$ and a uniform value type
\[
\text{Value} = \mathbb{N} \;\dot\cup\; \text{String} \;\dot\cup\; \text{Role}
\]
that absorbs each attribute payload. Column projections $\text{getVal}_J(\text{col}, j)$ and $\text{getVal}_S(\text{col}, s)$ extract the corresponding wrapped value from a record or a tuple.

\paragraph{Conditions.}
A condition $C$ is built from constants, atomic comparisons over a column, and Boolean combinations $\land,\lor$. It is interpreted over both formats:
\begin{itemize}
    \item $\text{eval}_J(C, j)$: evaluates $C$ on a JSON record $j$ via field access on the record;
    \item $\text{eval}_S(C, s)$: evaluates $C$ on an SQL tuple $s$ via projection of the tuple components.
\end{itemize}

\begin{lemma}[Condition Bridge]
\label{lem:cond-bridge}
For every condition $C$ and every $j \in \mathcal{J}$,
\[
\text{eval}_J(C, j) \;=\; \text{eval}_S\big(C,\ \text{toS}(j)\big).
\]
Equivalently, for every $s \in \mathcal{S}$, $\text{eval}_S(C, s) = \text{eval}_J(C, \text{toJ}(s))$.
\end{lemma}

\noindent The proof is by induction on $C$, using the column-projection bridge $\text{getVal}_J(\text{col}, j) = \text{getVal}_S(\text{col}, \text{toS}(j))$ established by case analysis on $\text{col}$.

\paragraph{Per-row update.}
For per-row mutation we define a typed but \emph{permissive} updater
\[
\text{applyUpdate}_J : \text{Col} \times \text{Value} \times \mathcal{J} \to \mathcal{J},
\qquad
\text{applyUpdate}_S : \text{Col} \times \text{Value} \times \mathcal{S} \to \mathcal{S},
\]
where the value's variant must agree with the column's payload type for the update to take effect; otherwise the row passes through unchanged. The conditional version $\text{applyUpdateIf}_J(\text{col}, v, C, j)$ updates iff $\text{eval}_J(C, j) = \text{true}$, and analogously on $\mathcal{S}$.

\begin{lemma}[Update Bridge]
\label{lem:update-bridge}
For every column $\text{col}$, value $v$, condition $C$, and record $j \in \mathcal{J}$,
\[
\text{applyUpdateIf}_S\big(\text{col}, v, C, \text{toS}(j)\big)
\;=\;
\text{toS}\big(\text{applyUpdateIf}_J(\text{col}, v, C, j)\big).
\]
\end{lemma}

\noindent The condition guard is shared via Lemma~\ref{lem:cond-bridge}; the unconditional core requires only case analysis on $(\text{col}, v)$ pairs.

% =====================================================

\subsection*{4. Query Languages and Translation}

The query language has six constructors. Five return a database; one returns a scalar. The translation $T : Q_{jq} \to Q_{sql}$ is structure-preserving:

\begin{center}
\begin{tabular}{lll}
\hline
jq constructor & SQL constructor & Result kind\\
\hline
$\texttt{find}\,(\text{col}, C)$        & $\texttt{SELECT}\,(\text{col}, C)$  & database \\
$\texttt{drop}\,(C)$                    & $\texttt{DELETE}\,(C)$              & database \\
$\texttt{prepend}\,(j)$                 & $\texttt{INSERT}\,(\text{toS}(j))$  & database \\
$\texttt{clear}$                        & $\texttt{TRUNCATE}$                 & database \\
$\texttt{count}$                        & $\texttt{COUNT}$                    & scalar ($\mathbb{N}$) \\
$\texttt{modify}\,(\text{col}, v, C)$   & $\texttt{UPDATE}\,(\text{col}, v, C)$ & database \\
\hline
\end{tabular}
\end{center}

\paragraph{Result type.}
Both evaluators return into a tagged sum
\[
\text{Result} = \mathcal{D} \;\dot\cup\; \mathbb{N},
\]
with $\mathcal{D} = \mathcal{JD}$ on the JSON side and $\mathcal{D} = \mathcal{SD}$ on the SQL side. Result equivalence dispatches by tag: two database results compare under the appropriate equivalence (set or permutation), two scalar results compare with $=$, and a tag mismatch is $\bot$.

% =====================================================

\subsection*{5. Query Semantics}

Writing $r = \text{toRows}(SD)$ throughout, and letting
\[
r_C = \{\, s \in r \mid \text{eval}_S(C, s) = \text{true} \,\},
\quad
r_{\lnot C} = \{\, s \in r \mid \text{eval}_S(C, s) = \text{false} \,\}.
\]

\paragraph{Filtering.}
\[
\text{eval}_{jq}(JD,\ \texttt{find}\,(\text{col}, C)) = \{\, j \in JD \mid \text{eval}_J(C, j) \,\},
\]
\[
\text{eval}_{jq}(JD,\ \texttt{drop}\,C) = \{\, j \in JD \mid \neg\text{eval}_J(C, j) \,\}.
\]
On the SQL side, the filter is taken on rows and then the columns are rebuilt:
\[
\text{eval}_{sql}(SD,\,\texttt{SELECT}\,(\text{col},C)) = \big(\,\pi_1^{*}(r_C),\ \pi_{2,1}^{*}(r_C),\ \pi_{2,2}^{*}(r_C)\,\big),
\]
and analogously for $\texttt{DELETE}$ with $r_{\lnot C}$. Here $\pi_1^{*}$, $\pi_{2,1}^{*}$, $\pi_{2,2}^{*}$ denote the pointwise projections to the three components.

\paragraph{Insertion and emptying.}
\[
\text{eval}_{jq}(JD,\ \texttt{prepend}\,j) = j :: JD,
\qquad
\text{eval}_{jq}(JD,\ \texttt{clear}) = [\,].
\]
\[
\text{eval}_{sql}(SD,\,\texttt{INSERT}\,s) = \big(\,s.1 :: SD.\text{names},\ s.2.1 :: SD.\text{ages},\ s.2.2 :: SD.\text{roles}\,\big),
\]
\[
\text{eval}_{sql}(SD,\,\texttt{TRUNCATE}) = (\,[\,],\ [\,],\ [\,]\,).
\]

\paragraph{Aggregate.}
\[
\text{eval}_{jq}(JD,\,\texttt{count}) = |JD|,
\qquad
\text{eval}_{sql}(SD,\,\texttt{COUNT}) = |\text{toRows}(SD)|.
\]
Going through $\text{toRows}$ rather than $|SD.\text{names}|$ ensures the count respects the columnar discipline: only matched-up triples are rows.

\paragraph{Per-row update.}
\[
\text{eval}_{jq}(JD,\,\texttt{modify}(\text{col},v,C)) = JD\,\textsf{.map}\,\text{applyUpdateIf}_J(\text{col},v,C),
\]
\[
\text{eval}_{sql}(SD,\,\texttt{UPDATE}(\text{col},v,C)) = \text{rebuild}\big(\,r\,\textsf{.map}\,\text{applyUpdateIf}_S(\text{col},v,C)\,\big),
\]
where $\text{rebuild}$ projects a row list back into three parallel columns.

\paragraph{Round-trip lemmas.}
The decomposition--transform--rebuild discipline used by every database-returning SQL query is sound:

\begin{lemma}[Filter Round-trip]
\label{lem:rt-filter}
For every $SD$ and predicate $p$,
\[
\text{toRows}\big(\text{rebuild}(r_p)\big) = r_p,
\quad\text{where } r_p = \{\,s \in \text{toRows}(SD) \mid p(s)\,\}.
\]
\end{lemma}

\begin{lemma}[Map Round-trip]
\label{lem:rt-map}
For every $SD$ and $f : \mathcal{S} \to \mathcal{S}$,
\[
\text{toRows}\big(\text{rebuild}\,(\text{toRows}(SD)\,\textsf{.map}\,f)\big) = \text{toRows}(SD)\,\textsf{.map}\,f.
\]
\end{lemma}

\begin{lemma}[Insert Round-trip]
\label{lem:rt-insert}
For every triple $s \in \mathcal{S}$ and $SD$,
\[
\text{toRows}\big(\,s.1 :: SD.\text{names},\ s.2.1 :: SD.\text{ages},\ s.2.2 :: SD.\text{roles}\,\big) = s :: \text{toRows}(SD).
\]
\end{lemma}

\noindent All three reduce to the identity $\text{zip}_3(\ell\,\textsf{.map}\,\pi_1,\ \ell\,\textsf{.map}\,\pi_{2,1},\ \ell\,\textsf{.map}\,\pi_{2,2}) = \ell$, by induction on the row list $\ell$. Together they show that the columnar layout is invisible at the row level.

% =====================================================

\subsection*{6. Main Correctness Theorem}

\begin{theorem}[Translation Correctness]
\label{thm:correctness}
Let $JD \in \mathcal{JD}$ and $SD \in \mathcal{SD}$ with $JD \approx SD$. For any jq query $q$,
\[
\text{eval}_{jq}(JD,\,q)
\;\equiv_R\;
\text{eval}_{sql}\big(SD,\,T(q)\big),
\]
where $\equiv_R$ is the result equivalence dispatching to $\equiv$ on database results and to $=$ on scalar results.
\end{theorem}

\begin{proof}
The hypothesis is permutation equivalence; for the database-returning cases, set equivalence is recovered via:

\begin{lemma}[Permutation implies set]
\label{lem:perm-set}
$JD \approx SD \implies JD \equiv SD$.
\end{lemma}

\noindent Membership in $JD\,\textsf{.map}\,\text{toS}$ is preserved by permutations, and membership in $JD$ is recovered by inverting $\text{toS}$ via $\text{toJ}$.

We split on the constructor of $q$.

\smallskip\noindent\emph{Filter cases (\texttt{find}, \texttt{drop}).}
For any $j \in \mathcal{J}$,
\[
j \in \text{eval}_{jq}(JD,\,\texttt{find}\,C)
\iff j \in JD \,\land\, \text{eval}_J(C, j).
\]
By Lemma~\ref{lem:perm-set}, $j \in JD \iff \text{toS}(j) \in r$; by Lemma~\ref{lem:cond-bridge}, the conditions agree pointwise; by Lemma~\ref{lem:rt-filter}, $r_C$ is exactly the row projection of the SQL output. Hence
\[
j \in \text{eval}_{jq}(JD,\,\texttt{find}\,C) \iff \text{toS}(j) \in \text{toRows}(\text{eval}_{sql}(SD,\,\texttt{SELECT}\,(\text{col},C))).
\]
The second clause of $\equiv$ is symmetric, applied through $\text{toJ}$. The \texttt{drop} case is identical with the predicate negated.

\smallskip\noindent\emph{Insert case (\texttt{prepend}).}
By Lemma~\ref{lem:rt-insert}, the SQL output's rows are $\text{toS}(j) :: r$. Membership reduces case-wise: a record equals $j$ (matched on both sides via $\text{toS}\,$\textendash$\,\text{toJ}$ inversion), or is in the underlying database (handled by Lemma~\ref{lem:perm-set}).

\smallskip\noindent\emph{Empty case (\texttt{clear}).}
Both sides reduce to the empty membership predicate: $u \in [\,] \iff \bot$ on the JSON side, $\text{toS}(u) \in \text{zip}_3([\,],[\,],[\,]) \iff \bot$ on the SQL side.

\smallskip\noindent\emph{Aggregate case (\texttt{count}).}
This case requires the strength of permutation equivalence: $|JD\,\textsf{.map}\,\text{toS}| = |\text{toRows}(SD)|$ by length-preservation under permutation, and $|JD\,\textsf{.map}\,\text{toS}| = |JD|$ since map preserves length.

\smallskip\noindent\emph{Update case (\texttt{modify}).}
This case also requires permutation equivalence. We chain three steps:
\begin{enumerate}
    \item Bridge equality (Lemma~\ref{lem:update-bridge}, applied pointwise via $\text{map}\circ\text{map}$):
    \[
    (JD\,\textsf{.map}\,\text{applyUpdateIf}_J(\dots))\,\textsf{.map}\,\text{toS}
    \;=\; (JD\,\textsf{.map}\,\text{toS})\,\textsf{.map}\,\text{applyUpdateIf}_S(\dots).
    \]
    \item Permutation transport: from $JD\,\textsf{.map}\,\text{toS} \approx \text{toRows}(SD)$ and the fact that $\textsf{.map}\,f$ preserves permutations,
    \[
    (JD\,\textsf{.map}\,\text{toS})\,\textsf{.map}\,\text{applyUpdateIf}_S(\dots) \;\approx\; \text{toRows}(SD)\,\textsf{.map}\,\text{applyUpdateIf}_S(\dots).
    \]
    \item Map round-trip (Lemma~\ref{lem:rt-map}): the right-hand side equals the row projection of the SQL output.
\end{enumerate}
This establishes permutation equivalence between the updated JSON list and the SQL output's rows; Lemma~\ref{lem:perm-set} lifts it to the required set equivalence.
\end{proof}

% =====================================================

\subsection*{7. Discussion}

The theorem covers a query language that is not just filtering: it includes order-sensitive insertion, an aggregate, and per-row mutation. The proof factors through three reusable ingredients.

\begin{itemize}
    \item The \emph{Condition Bridge Lemma} keeps predicate evaluation in lockstep across $\mathcal{J}$ and $\mathcal{S}$, letting filter and update reason about the same Boolean function viewed through either schema.
    \item The \emph{Update Bridge Lemma} extends that lockstep to per-row mutation: $\text{toS}$ commutes with $\text{applyUpdateIf}$, so JSON-side and SQL-side updates produce corresponding rows.
    \item The \emph{round-trip lemmas} (filter, map, insert) make the columnar storage discipline transparent. Every database-returning SQL query factors through a row list; the round-trip lemmas show that the rebuilt $\mathcal{SD}$ recovers exactly that list under $\text{toRows}$, so every claim about row-level behaviour lifts to the columnar representation.
\end{itemize}

The aggregate case is the only one that genuinely needs permutation equivalence rather than the weaker set equivalence: \texttt{count} depends on multiplicity. For all five database-returning constructors, set equivalence suffices, and the precondition $JD \approx SD$ is exploited only through Lemma~\ref{lem:perm-set}. This stratification is exactly what one would expect: the moment a query observes the cardinality of the database, multiplicities must be tracked.

\end{document}